---
title: "Thinking through a power analysis for a case series method for deaths and high heat days"
author: "Brooke Anderson"
date: "September 2, 2015"
output: html_document
---

```{r, message = FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
```

## Simple example

First, I'll need to decide how rare the event is. Let's say that only 2% of days are "high heat days". Now I'll take a warm season of data and randomly put in high heat days on 5% of those days:

```{r}
sim_data <- data.frame(date = seq(from = as.Date("2000-01-01"),
                to = as.Date("2009-12-31"), by = 1)) %>%
        filter(months(date) %in% c("May", "June", "July",
                                  "August", "September"))
table(months(sim_data$date))
```

```{r}
heat_days <- sample(sim_data$date, round(nrow(sim_data) * 0.02))
sim_data <- mutate(sim_data, 
                   heat_day = date %in% heat_days,
                   year = format(date, "%Y"),
                   day_of_year = yday(date))
head(sim_data, 20)
```

The following plot shows the days that were randomly selected as heat days:

```{r}
ggplot(sim_data, aes(x = day_of_year, y = heat_day)) + 
        geom_point() + 
        facet_grid(year ~ . )
```

The next thing I'll need to decide are: 

1. What is the baseline rate of the outcome (e.g., cardiovascular death)
2. What is the effect size of a heat day (i.e., the relative incidence for heat days versus other days)

I'll check out what I can see if I set the baseline rate of the outcome to be 50 per day and the relative incidence for a heat day to be 1.50 (i.e., a 50% increase):

```{r}
baseline_rate <- 50
RI <- 1.50
```

Now I'll simulate the numbers of outcomes on each day. I'm assuming a Poisson distribution for the number of deaths:

```{r}
sim_data$outcome <- NA
risk_numbers <- table(sim_data$heat_day)
sim_data$outcome[sim_data$heat_day == FALSE] <- rpois(
                                                risk_numbers["FALSE"],
                                                       baseline_rate)
sim_data$outcome[sim_data$heat_day == TRUE] <- rpois(
                                                risk_numbers["TRUE"],
                                                baseline_rate * RI)
```

```{r}
ggplot(sim_data, aes(x = day_of_year, y = outcome, color = heat_day)) +
        geom_point() + 
        facet_grid(year ~ .)
```

You can see from this that, with a relative incidence of 1.50 and all the other assumptions, you can clearly pick out an increase in incidence during risk days just by eye-balling the figure.

Now, I can run a GLM on this to see if I identify the effect:

```{r}
mod_1 <- glm(outcome ~ heat_day, data = sim_data, 
             family = poisson(link = "log"))
summary(mod_1)$coef
exp(summary(mod_1)$coef[2, 1])
exp(confint(mod_1)[2, ]) 
```

This has generated a point estimate almost identical to the RI used in the model, and the confidence interval is surrounding the specified RI. The p-value for the heat days coefficient is `r summary(mod_1)$coef[2, 4]`, well below 0.05.

## Simple function to estimate power

Next, I'll take these ideas and put them into a simple function that takes in inputs of number of study days, percent of days in the high risk category, assumed effect size for being in a risk period (relative incidence), and the baseline rate of the outcome. The function will do a similar calculation to what I just did, and then will result in a yes / no answer of whether, for a single simulation, it found that the p-value of the effect size was smaller than 0.05. I can then run this function in `replicate()` to simulate many times and come up with the proportion of the time that the null hypothesis failed to be rejected at a true effect size of that specified by the RI value.